\noindent
{\bf IMPORTANT NOTICE}

{\small
  This documentation is furnished for informational use only
  and is subject to change without notice. Galois, Inc.~assumes no
  responsibility or liability for any errors or inaccuracies that may
  appear in this documentation. Of course, we appreciate bug reports
  and clarification suggestions.

  Copyright 2003--2015 Galois, Inc. All rights reserved by Galois,
  Inc.

  The software installed in accordance with this documentation is
  copyrighted and licensed by Galois, Inc.~under separate license
  agreement. This software may only be used pursuant to the terms and
  conditions of such license agreement.}

\noindent
{\bf TRADEMARKS}

{\small 
  Cryptol is a registered trademark of Galois, Inc.~in the United
  States and other countries. UNIX is a registered trademark of The
  Open Group in the U. S.  and other countries. Linux is a registered
  trademark of Linus Torvalds.

  Other company or product names mentioned herein may be trademarks or
  registered trademarks of their respective owners. Trademark
  specifications are subject to change without notice. All terms
  mentioned in this documentation that are known to be trademarks or
  service marks have been appropriately capitalized to the best of our
  knowledge; however, Galois cannot attest to the accuracy of all
  trademark information. Use of a term in this documentation should
  not be regarded as affecting the validity of any trademark or
  service mark.

\vspace{0.5in}
\noindent
Galois, Inc.\\
421 SW Sixth Avenue, Suite 300 \\
Portland, OR 97204
}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "../main/Cryptol"
%%% End: 

